---
title: "Rabies Outbreak MIMAROPA — Solutions"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(reactable)
library(scales)

province_names <- c("Romblon", "Oriental Mindoro", "Marinduque", "Occidental Mindoro")

## load outbreak data
dat_outbreak <- read_csv(here("data","epi","dat_all_anon.csv"))
dat_outbreak <- dat_outbreak %>% mutate(date = dmy(date))
mdv_summary <- read.csv(here("data", "epi", "mdv_summary.csv"))

## Filtering outbreak data
dat_outbreak_trun <- dat_outbreak %>% 
  filter(date >= "2023-06-30" & date < "2024-12-31") 

## Geographic data for mapping
municipality <- st_read(here("data", "gis"), layer = "PHL_municipality", quiet = TRUE)
village_loc  <- read.csv(here("data/gis", "PHL_village_centroids.csv"))

# Colour palettes used throughout
pal_status <- c("Human Death"        = "#A3003F",
                "Probable (Animal)"  = "#E59500",
                "Confirmed (Animal)" = "#3C8590")

pal_type   <- c("Human" = "#A3003F",
                "Animal" = "#3C8590")

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_outbreak$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak$date), "month"),
  by   = "month"
) - 1

monthly_breaks_trun <- seq.Date(
  from = floor_date(min(dat_outbreak_trun$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak_trun$date), "month"),
  by   = "month"
) - 1
```

---

## Overview

This document works through a series of exploratory plots that help characterise
the rabies outbreak across MIMAROPA in Romblon, Oriental Mindoro, Marinduque, and Occidental
Mindoro. Code is folded but available via the **Show code** button.

---

## 1. How did the outbreak evolve over time?

### 1.1 Epidemic curve 

Monthly reported animal cases (confirmed and probable) and human deaths in Oriental Mindoro, Romblon, Marinduque and Occidental Mindoro

```{r epi-curve-overall, fig.width=7, fig.height=4}

monthly_breaks_trun <- seq.Date(
  from = floor_date(min(dat_outbreak_trun$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak_trun$date), "month"),
  by   = "month"
) - 1

ggplot(dat_outbreak_trun, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks_trun, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "3 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top")
```

### 1.2 Location of cases

Spatial distribution of all cases. Human deaths are
shown as triangles, whilst animal cases are shown as circles.

```{r map, fig.width=6, fig.height=6}


# Attach coordinates
village_loc_filt <- village_loc %>%
  filter(str_starts(Loc_ID, paste(paste0(province_names, "-"), collapse = "|")))

dat_mapped <- dat_outbreak_trun %>%
  mutate(loc_id = paste(province, municipality, barangay, sep = "-")) %>%
  left_join(select(village_loc_filt, Loc_ID, Latitude, Longitude),
            by = c("loc_id" = "Loc_ID"))

municipality_df <- municipality %>%
  filter(NAME_1 %in% province_names)

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.6)) +
  labs(colour = "", shape = "", alpha = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position  = c(0.85, 0.85),
        panel.grid.major = element_blank(),
        axis.text        = element_blank(),
        axis.title       = element_blank())
```



> **Next steps:**
>
> - Explore the whole outbreak period (2019 - 2026)?
> - What does looking at this data by province tell us? 
> - What specific challenges might have been encountered in each province?
> - What can you say about the overall outbreak? Where should we focus our surveillance and intervention efforts?
> - What additional plots and information would you want to be able to understand this better?


<details>
<summary><strong>💡 Hints</strong></summary>

- These plots combines all provinces — try breaking this plots down by province. For Romblon it may be useful to break this down further by island (Tablas, Sibuyan, and Corcuera).
- Try zooming in on a particular time period (e.g. 2019–2021).
- Roughly 1 person is bitten every 3 rabid dogs. Of people bitten, 90% get PEP. Of the 10% that don't get PEP, 20% will develop rabies and die. How does this change your interpretation of the early outbreak period?
- How do the population sizes above change your interpretation of the case counts? 

</details>

---

## 2 How did the outbreak evolve in each province?

Looking at the epidemic curve by province reveals whether provinces were affected
simultaneously or sequentially — important for understanding transmission routes and introduction events

### 2.1 Oriental Mindoro

> **Key questions and things to explore:**
>
> - What is happenning in the early outbreak period? 
> - How do animal cases and human cases compare in Oriental Mindoro
> - Roughly 1 person is bitten every 3 rabid dogs. Of people bitten, 90% get PEP. Of the 10% that don't get PEP, 20% will develop rabies and die. How does this change your interpretation of the early outbreak period?
> - How does the human in April 2026 change your interpretation of what is happenning in Ormin 

```{r epi-curve-ormin, fig.width=9, fig.height=3.5}

# Filter MDV data for Oriental Mindoro and create date column
mdv_ormin <- mdv_summary %>%
  filter(region == "Ormin") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

dat_outbreak %>%
  filter(province == "Oriental Mindoro") %>%
  ggplot(aes(x = date, fill = case_status_fct)) +
  geom_rect(
    data = mdv_ormin,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(
    values = c(pal_status)
  ) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 40, vjust = 0.5),
    legend.position  = "top",
    strip.background = element_rect(fill = "#f0f0f0"),
    strip.text       = element_text(face = "bold")
  )

```
*Dates of mass dog vaccination campaigns are shown in grey (MDV campaign defined as >700 dogs vaccinated in a municipality in a month)*

#### How do human cases compare to animal cases in Ormin?

```{r epi-curve-facet-ormin, fig.width=9, fig.height=5}

dat_outbreak %>% 
  filter(province == "Oriental Mindoro") %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~case_type, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

#### Distribution of cases in Ormin over time

```{r map-ormin, fig.width=11, fig.height=7}


dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Oriental Mindoro") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

municipality_df <- municipality %>%
  filter(NAME_1 == "Oriental Mindoro")

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year,ncol=4) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

> **Key questions:**

### 2.2. Romblon

> **Key questions:**
>
> - How did the outbreak evolve in Romblon over time? Romblon is made up on different islands, how did it spread over time?
> - What do these plots tell you about the current surveillance efforts in Romblon?
> - When did the first death occur? And how does this correspond to the increase in dog vaccination efforts? 


#### How does the outbreak look in Romblon if we break this down by island?

```{r epi-curve-romblon, fig.width=9, fig.height=7}

dat_romblon = dat_outbreak %>% 
  filter(province == "Romblon")

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_romblon$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_romblon$date), "month"),
  by   = "month"
) - 1

mdv_romblon <- mdv_summary %>%
  filter(region == "Romblon") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

dat_romblon %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
    geom_rect(
    data = mdv_romblon,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```
*Dates of mass dog vaccination campaigns are shown in grey (MDV campaign defined as >700 dogs vaccinated in a municipality in a month)*


```{r epi-curve-island, fig.width=9, fig.height=7}


dat_romblon %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~island, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

#### Spatial distributions of cases in Romblon by year

```{r map-romblon, fig.width=9, fig.height=7}


dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Romblon") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

municipality_df <- municipality %>%
  filter(NAME_1 == "Romblon")

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```


### 2.3 Marinduque

> **Key questions and things to consider:**
>
> - How did the outbreak evolve in Marinduque over time?
> - What do these plots tell you about the current surveillance efforts in Marinduque?
> - When did the first death occur? And how does this correspond to the increase in dog vaccination efforts? 
> - How does the outbreak in Marinduque compare to the outbreaks in Ormin and Romblon?
> - How does the human death in Dec 2025 impact your interpretation


```{r epi-curve-marin, fig.width=9, fig.height=4}

# Filter MDV data for Marinduque and create date column
mdv_marin <- mdv_summary %>%
  filter(region == "Marinduque") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

dat_marin = dat_outbreak %>% 
  filter(province == "Marinduque")

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_marin$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_marin$date), "month"),
  by   = "month"
) - 1

dat_outbreak %>% 
  filter(province == "Marinduque") %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_rect(
    data = mdv_marin,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~island, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```
*Dates of mass dog vaccination campaigns are shown in grey (MDV campaign defined as >700 dogs vaccinated in a municipality in a month)*



```{r map-marin, fig.width=10, fig.height=6}


dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Marinduque") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

municipality_df <- municipality %>%
  filter(NAME_1 == "Marinduque")

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

---

## 3. Providing the regional context

> **Key questions and things to explore:**
>
> - Which province do you think has the most robust surveillance? 
> - Do you consider rabies to be endemic in any of the provinces? 
> - What do you think are the triggers of outbreak responses?
> - How to the ratios of human to animal cases differ in the three provinces?

### 3.1 Extended epidemic curve 

Monthly reported animal cases (confirmed and probable) and human deaths in Oriental Mindoro, Romblon, Marinduque and Occidental Mindoro including all data from 2019 - 2026

```{r epi-curve-region, fig.width=9, fig.height=3.5}

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_outbreak$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak$date), "month"),
  by   = "month"
) - 1

ggplot(dat_outbreak, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "3 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top")
```

### 3.2 Epidemic curve by province

```{r epi-curve-facet, fig.width=9, fig.height=8}
ggplot(dat_outbreak, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~province, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "",
       title = "Reported rabies cases by province") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

### 3.3 Regional location of cases

Spatial distribution of all cases from 2019 to 2026. Human deaths are
shown as triangles, whilst animal cases are shown as circles.

```{r map-region, fig.width=9, fig.height=9}

# Attach coordinates
village_loc_filt <- village_loc %>%
  filter(str_starts(Loc_ID, paste(paste0(province_names, "-"), collapse = "|")))

dat_mapped <- dat_outbreak %>%
  mutate(loc_id = paste(province, municipality, barangay, sep = "-")) %>%
  left_join(select(village_loc_filt, Loc_ID, Latitude, Longitude),
            by = c("loc_id" = "Loc_ID"))

municipality_df <- municipality %>%
  filter(NAME_1 %in% province_names) 

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
    facet_wrap(~ year) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.7)) +
  labs(colour = "", shape = "", alpha = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        panel.grid.major = element_blank(),
        axis.text        = element_blank(),
        axis.title       = element_blank())



```

### 3.4. How do human deaths compare with animal cases?

A critical question is whether surveillance is sufficient. If we have areas where human deaths occur 
with relatively few reported animal cases this would suggest under-reporting
of animal cases. Therefore, the incidence of the ratio of human deaths to animal cases may provide an estimate of the surveillance sensitivity and reporting.

#### Population estimates by province

| Location | Human population | Dog population |
|---|--:|--:|
| Oriental Mindoro | 900,000 | 200,000 |
| Occidental Mindoro | 510,000 | 110,000 |
| Marinduque | 220,000 | 40,000 |
| Romblon (Tablas) | 170,000 | 30,000 |
| Romblon (Sibuyan) | 65,000 | 15,000 |
| Romblon (Corcuera) | 9,000 | 2,000 |


#### 3.4.1 Incidence of human deaths and animal cases

```{r incidence-table}
# Population data
dog_pop <- c(
  "Oriental Mindoro"    = 200000,
  "Occidental Mindoro"    = 110000,
  "Marinduque" = 40000,
  "Tablas Island (Romblon)"     = 30000,
  "Sibuyan Island (Romblon)"    = 15000,
  "Corcuera Island (Romblon)"   = 2000,
  "Romblon Island (Romblon)" = 8800
  
)

human_pop <- c(
  "Oriental Mindoro"    = 920000,
    "Occidental Mindoro"    = 510000,
  "Marinduque" = 220000,
  "Tablas Island (Romblon)"     = 170000,
  "Sibuyan Island (Romblon)"    = 65000,
  "Corcuera Island (Romblon)"   = 9000,
  "Romblon Island (Romblon)" = 40000
)

dat_outbreak %>%
  filter(!is.na(island)) %>%                          # drop unmatched locations
  mutate(years = as.numeric(max(date, na.rm = TRUE) - 
                            min(date, na.rm = TRUE)) / 365.25) %>% 
  group_by(island) %>%
  summarise(
    animal_cases = sum(case_type == "Animal"),
    human_deaths = sum(case_type == "Human"),
    years        = first(years),
    dogs         = dog_pop[first(island)],
    humans       = human_pop[first(island)],
    .groups = "drop"
  ) %>%
  mutate(
    animal_incidence = round(animal_cases / (dogs * years) * 1000, 2),
    human_incidence  = round(human_deaths / (humans * years) * 100000, 2)
  ) %>%
  select(
    Island            = island,
    `Animal cases`    = animal_cases,
    `Animal incidence (per 1,000 dogs/year)` = animal_incidence,
    `Human deaths`    = human_deaths,
    `Human incidence (per 100,000 people/year)` = human_incidence
  ) %>%
  reactable(striped = TRUE, highlight = TRUE, pagination = FALSE)
```

#### 3.4.2 Ratio of human deaths to animal cases
```{r ratio-table}
dat_outbreak %>%
  group_by(province) %>%
  summarise(
    Animal_cases  = sum(case_type == "Animal"),
    Human_deaths  = sum(case_type == "Human"),
    .groups = "drop"
  ) %>%
  mutate(Deaths_per_100_animal = round(Human_deaths / Animal_cases * 100, 1)) %>%
  rename(Province               = province,
         `Animal cases`         = Animal_cases,
         `Human deaths`         = Human_deaths,
         `Deaths per 100 animal cases` = Deaths_per_100_animal) %>%
  reactable(striped = TRUE, highlight = TRUE, pagination = FALSE)
```





