---
title: "Responding to a Zoonotic Disease Outbreak - Epidemiological interpretation and implications for decision-making in One Health"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(reactable)
library(scales)

province_names <- c("Romblon", "Oriental Mindoro", "Marinduque", "Occidental Mindoro")

## load outbreak data
dat_outbreak <- read_csv(here("data","epi","dat_all_anon.csv"))
dat_outbreak <- dat_outbreak %>% mutate(date = dmy(date))
mdv_summary <- read.csv(here("data", "epi", "mdv_summary.csv"))

## Filtering outbreak data
dat_outbreak_trun <- dat_outbreak %>% 
  filter(date >= "2023-06-30" & date < "2024-12-31") 

## Geographic data for mapping
municipality <- st_read(here("data", "gis"), layer = "PHL_municipality", quiet = TRUE)
village_loc  <- read.csv(here("data/gis", "PHL_village_centroids.csv"))

# Colour palettes used throughout
pal_status <- c("Human Death"        = "#A3003F",
                "Probable (Animal)"  = "#E59500",
                "Confirmed (Animal)" = "#3C8590")

pal_type   <- c("Human" = "#A3003F",
                "Animal" = "#3C8590")

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_outbreak$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak$date), "month"),
  by   = "month"
) - 1

monthly_breaks_trun <- seq.Date(
  from = floor_date(min(dat_outbreak_trun$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak_trun$date), "month"),
  by   = "month"
) - 1
```

---

## Overview

This document works through a series of exploratory plots that help characterise
a recent rabies outbreak across MIMAROPA region (spanning Romblon, Oriental Mindoro, Marinduque, and Occidental Mindoro provinces). 

Code is folded but available via the **Show code** button.

---

## 1. Describing the outbreak:

### 1.1 Epidemic curve 

Monthly reported animal rabies cases (confirmed and probable) and human rabies deaths in Oriental Mindoro, Romblon, Marinduque and Occidental Mindoro are presented in a time series.

```{r epi-curve-overall, fig.width=7, fig.height=4}

monthly_breaks_trun <- seq.Date(
  from = floor_date(min(dat_outbreak_trun$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak_trun$date), "month"),
  by   = "month"
) - 1

ggplot(dat_outbreak_trun, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks_trun, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "3 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top")
```

### 1.2 Locations of cases

The spatial distribution of all these cases are mapped. Human deaths are
shown as triangles, whilst animal cases are shown as circles.

```{r map, fig.width=6, fig.height=6}


# Attach coordinates
village_loc_filt <- village_loc %>%
  filter(str_starts(Loc_ID, paste(paste0(province_names, "-"), collapse = "|")))

dat_mapped <- dat_outbreak_trun %>%
  mutate(loc_id = paste(province, municipality, barangay, sep = "-")) %>%
  left_join(select(village_loc_filt, Loc_ID, Latitude, Longitude),
            by = c("loc_id" = "Loc_ID"))

municipality_df <- municipality %>%
  filter(NAME_1 %in% province_names)

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.6)) +
  labs(colour = "", shape = "", alpha = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position  = c(0.85, 0.85),
        panel.grid.major = element_blank(),
        axis.text        = element_blank(),
        axis.title       = element_blank())
```



> **Next steps:**
>
> - How does your interpretation of the epidemiological situation change when you view the whole time period (2019 - 2026)?
> - What does looking at these data by province tell us? 
> - What specific challenges might have been encountered in each province?
> - What can you say about the overall epidemiological situation? 
Which provinces suffered the worst outbreaks and which are now closest to having rabies under control and may even be approaching elimination?
> - Where should surveillance and intervention efforts be strengthened?
> - What additional plots and information would you want to be able to understand this better?


<details>
<summary><strong>💡 Hints</strong></summary>

- These plots combines all provinces — try breaking this plot down by province. For Romblon it may be useful to break this down further by island (Tablas, Sibuyan, and Corcuera).
- Try zooming in on a particular time period (e.g. 2019–2021).
- How helpful is the human surveillance data without the animal surveillance? and vice versa? with both sets of data combined what new conclusions do you draw?
- How do the population sizes above change your interpretation of the case counts? 

</details>

---

## 2. How did the epidemiological situation evolve in each province?

Looking at the epidemic curve by province reveals whether provinces were affected
simultaneously or sequentially — this is important for understanding routes of disease spread and the timing of introduction events

### 2.1 Oriental Mindoro

> **Key questions and things to explore:**
>
> - What is happening in the early outbreak period? 
> - How do animal cases compare to human cases in Oriental Mindoro?
> - Roughly 1 person is bitten every 3 rabid dogs. Of people bitten, 90% get PEP. Of the 10% that don't get PEP, 20% will develop rabies and die. How does this change your interpretation of the early outbreak period?
> - How does the most recent human death in April 2026 (look at the 2026 map! **should we not show the very end of the timeseries?**) change your interpretation of what is happening in Oriental Mindoro? 

```{r epi-curve-ormin, fig.width=9, fig.height=3.5}

# Filter MDV data for Oriental Mindoro and create date column
mdv_ormin <- mdv_summary %>%
  filter(region == "Ormin") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

dat_outbreak %>%
  filter(province == "Oriental Mindoro") %>%
  ggplot(aes(x = date, fill = case_status_fct)) +
  geom_rect(
    data = mdv_ormin,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(
    values = c(pal_status)
  ) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 40, vjust = 0.5),
    legend.position  = "top",
    strip.background = element_rect(fill = "#f0f0f0"),
    strip.text       = element_text(face = "bold")
  )

```
*Dates of mass dog vaccination campaigns are shown in grey (a campaign is defined as >700 dogs vaccinated in a municipality in a month)*

#### How do human cases compare to animal cases in Oriental Mindoro?

```{r epi-curve-facet-ormin, fig.width=9, fig.height=5}

dat_outbreak %>% 
  filter(province == "Oriental Mindoro") %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~case_type, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

#### Does the distribution of cases over time and across municipalities suggest any gaps in surveillance (from the health or animal health perspective)?

```{r map-ormin, fig.width=11, fig.height=7}


dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Oriental Mindoro") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

municipality_df <- municipality %>%
  filter(NAME_1 == "Oriental Mindoro")

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year,ncol=4) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```



### 2.2. Romblon

> **Key questions and things to consider:**
>
> - Romblon province is made up of different islands. How did rabies spread across the islands over time?
> - If you were facing the situation in Tablas island in December 2022 (look at the 2022 map), what actions would you recommend to the local authorities?
- When did the first death occur? What can you say about the responsiveness of dog vaccination efforts? 
> - What if you were facing the situation in Sibuyan island in December 2023, would you be concerned? what communications would you want between municipalities? How might your thinking change by August 2024? Do you have any concerns if you are from Corcuera island?
>  What do these complete plots tell you about the current surveillance efforts in Romblon?


#### How does the outbreak look in Romblon if we break this down by island?

```{r epi-curve-romblon, fig.width=9, fig.height=7}

dat_romblon = dat_outbreak %>% 
  filter(province == "Romblon")

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_romblon$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_romblon$date), "month"),
  by   = "month"
) - 1

mdv_romblon <- mdv_summary %>%
  filter(region == "Romblon") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

dat_romblon %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
    geom_rect(
    data = mdv_romblon,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```
*Dates of mass dog vaccination campaigns are shown in grey (MDV campaign defined as >700 dogs vaccinated in a municipality in a month)*


```{r epi-curve-island, fig.width=9, fig.height=7}


dat_romblon %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~island, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

#### Spatial distributions of human and animal rabies cases in Romblon by year

```{r map-romblon, fig.width=9, fig.height=7}


dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Romblon") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

municipality_df <- municipality %>%
  filter(NAME_1 == "Romblon")

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```


### 2.3 Marinduque

> **Key questions and things to consider:**
>
> - How did the outbreak evolve in Marinduque over time?
> - What do these plots tell you about the current surveillance efforts in Marinduque?
> - When did the first death occur? And how does this correspond to the increase in dog vaccination efforts? 
> - How does the outbreak in Marinduque compare to the outbreaks in Oriental mindoro and Romblon provinces?
> - How does the human death in Dec 2025 impact your interpretation?


```{r epi-curve-marin, fig.width=9, fig.height=4}

# Filter MDV data for Marinduque and create date column
mdv_marin <- mdv_summary %>%
  filter(region == "Marinduque") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

dat_marin = dat_outbreak %>% 
  filter(province == "Marinduque")

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_marin$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_marin$date), "month"),
  by   = "month"
) - 1

dat_outbreak %>% 
  filter(province == "Marinduque") %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_rect(
    data = mdv_marin,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~island, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```
*Dates of mass dog vaccination campaigns are shown in grey (a campaign defined as >700 dogs vaccinated in a municipality in a month)*



```{r map-marin, fig.width=10, fig.height=6}


dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Marinduque") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

municipality_df <- municipality %>%
  filter(NAME_1 == "Marinduque")

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

---

## 3. Providing the regional context

> **Key questions and things to explore:**
>
> - Which province do you think has the most robust surveillance? (consider surveillance from the veterinary perspective, the public health perspective and under a One Health approach) 
> - Do you consider rabies to be endemic in any of the provinces? 
> - What do you think are the triggers of outbreak responses?
> - Why do the ratios of human to animal cases differ in the three provinces?

### 3.1 Extended epidemic curve 

Monthly reported animal cases (confirmed and probable) and human deaths in Oriental Mindoro, Romblon, Marinduque and Occidental Mindoro including all data from 2019 - 2026

```{r epi-curve-region, fig.width=9, fig.height=3.5}

monthly_breaks <- seq.Date(
  from = floor_date(min(dat_outbreak$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak$date), "month"),
  by   = "month"
) - 1

ggplot(dat_outbreak, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "3 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top")
```

### 3.2 Epidemic curve by province

```{r epi-curve-facet, fig.width=9, fig.height=8}
ggplot(dat_outbreak, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~province, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "",
       title = "Reported rabies cases by province") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

### 3.3 Regional location of cases

Spatial distribution of all cases from 2019 to 2026. Human deaths are
shown as triangles, whilst animal cases are shown as circles.

```{r map-region, fig.width=9, fig.height=9}

# Attach coordinates
village_loc_filt <- village_loc %>%
  filter(str_starts(Loc_ID, paste(paste0(province_names, "-"), collapse = "|")))

dat_mapped <- dat_outbreak %>%
  mutate(loc_id = paste(province, municipality, barangay, sep = "-")) %>%
  left_join(select(village_loc_filt, Loc_ID, Latitude, Longitude),
            by = c("loc_id" = "Loc_ID"))

municipality_df <- municipality %>%
  filter(NAME_1 %in% province_names) 

ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
    facet_wrap(~ year) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.7)) +
  labs(colour = "", shape = "", alpha = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        panel.grid.major = element_blank(),
        axis.text        = element_blank(),
        axis.title       = element_blank())



```

### 3.4. How do human deaths compare with animal cases?

A critical question is whether surveillance is sufficient. If we have areas where human deaths occur 
with relatively few reported animal cases this would suggest under-reporting
of animal cases. Therefore, the incidence of the ratio of human deaths to animal cases may provide an estimate of the surveillance sensitivity and reporting.

#### Population estimates by province

| Location | Human population | Dog population |
|---|--:|--:|
| Oriental Mindoro | 900,000 | 200,000 |
| Occidental Mindoro | 510,000 | 110,000 |
| Marinduque | 220,000 | 40,000 |
| Romblon (Tablas) | 170,000 | 30,000 |
| Romblon (Sibuyan) | 65,000 | 15,000 |
| Romblon (Corcuera) | 9,000 | 2,000 |


#### 3.4.1 Incidence of human deaths and animal cases

```{r incidence-table}
# Population data
dog_pop <- c(
  "Oriental Mindoro"    = 200000,
  "Occidental Mindoro"    = 110000,
  "Marinduque" = 40000,
  "Tablas Island (Romblon)"     = 30000,
  "Sibuyan Island (Romblon)"    = 15000,
  "Corcuera Island (Romblon)"   = 2000,
  "Romblon Island (Romblon)" = 8800
  
)

human_pop <- c(
  "Oriental Mindoro"    = 920000,
    "Occidental Mindoro"    = 510000,
  "Marinduque" = 220000,
  "Tablas Island (Romblon)"     = 170000,
  "Sibuyan Island (Romblon)"    = 65000,
  "Corcuera Island (Romblon)"   = 9000,
  "Romblon Island (Romblon)" = 40000
)

dat_outbreak %>%
  filter(!is.na(island)) %>%                          # drop unmatched locations
  mutate(years = as.numeric(max(date, na.rm = TRUE) - 
                            min(date, na.rm = TRUE)) / 365.25) %>% 
  group_by(island) %>%
  summarise(
    animal_cases = sum(case_type == "Animal"),
    human_deaths = sum(case_type == "Human"),
    years        = first(years),
    dogs         = dog_pop[first(island)],
    humans       = human_pop[first(island)],
    .groups = "drop"
  ) %>%
  mutate(
    animal_incidence = round(animal_cases / (dogs * years) * 1000, 2),
    human_incidence  = round(human_deaths / (humans * years) * 100000, 2)
  ) %>%
  select(
    Island            = island,
    `Animal cases`    = animal_cases,
    `Animal incidence (per 1,000 dogs/year)` = animal_incidence,
    `Human deaths`    = human_deaths,
    `Human incidence (per 100,000 people/year)` = human_incidence
  ) %>%
  reactable(striped = TRUE, highlight = TRUE, pagination = FALSE)
```

#### 3.4.2 Ratio of human deaths to animal cases
```{r ratio-table}
dat_outbreak %>%
  group_by(province) %>%
  summarise(
    Animal_cases  = sum(case_type == "Animal"),
    Human_deaths  = sum(case_type == "Human"),
    .groups = "drop"
  ) %>%
  mutate(Deaths_per_100_animal = round(Human_deaths / Animal_cases * 100, 1)) %>%
  rename(Province               = province,
         `Animal cases`         = Animal_cases,
         `Human deaths`         = Human_deaths,
         `Deaths per 100 animal cases` = Deaths_per_100_animal) %>%
  reactable(striped = TRUE, highlight = TRUE, pagination = FALSE)
```

### 3.5. Is rabies an emerging infectious disease threat or a neglected disease of poverty?
Rabies can be viewed as an emerging disease that can cause outbreaks in areas where it was previously absent. But it is also a disease that is entirely vaccine-preventable, that can be eliminated through dog vaccination, whilst human deaths only occur if people at-risk do not access emergency life-saving vaccines after an exposure.

The last decade in MIMAROPA region illustrates all of these different situations. Can you quantify the incidence of human rabies deaths and of rabies in dogs in each province and how this varies by year?
What is the highest incidence? and what is the lowest incidence?
Taking these extremes, and extrapolating to the entirety of the Philippines, what is the worst case scenario? how many rabies deaths *could* occur across the country if they were all to face the same challenges in rabies control and prevention?

And what if rabies control improved and surveillance was better - what is the best case scenario? 
What are the obstacles to achieving this best case outcome? and what can be done better that can support the best outcome?!



