---
title: "Responding to a Zoonotic Disease Outbreak - Epidemiological interpretation and implications for decision-making in One Health"
subtitle: "Afternoon hands-on tutorial"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
    fig-width: 12
    fig-height: 6
    df-print: paged
    fontsize: 15px
    header-includes: |
      <style>
        body         { font-size: 15px; line-height: 1.5; }
        h1           { font-size: 1.7em; }
        h2           { font-size: 1.35em; }
        h3           { font-size: 1.15em; }
        h4           { font-size: 1.0em; }
        pre code     { font-size: 0.85em; }
        code         { font-size: 0.9em; }
        #TOC         { font-size: 13px; }
        .callout     { font-size: 0.95em; }
      </style>
  pdf:
    toc: true
    number-sections: true
    fig-width: 12
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
  
---

---

## Overview

We will explore surveillance data from Integrated Bite Case Management (IBCM) in MIMAROPA region.
IBCM is a One Health approach to rabies surveillance that links post-exposure vaccination of bite victims, investigation of biting animals, community sensitisation and programmatic coordination, to improve prevention, detection and response. 
IBCM was introduced to Romblon and Oriental Mindoro provinces in 2020, although COVID-19 restrictions prevented full implementation. In 2022 IBCM expanded to Marinduque province and in 2024 to Occidental Mindoro.

Using these data you will examine where outbreaks occurred; how these outbreaks spread; and why.
In groups you will look more deeply into one province and consider the implications for control and prevention.


## Learning objectives

In this session you will:

1) gain understanding of how to **critically interpret epidemiological data** considering a **One Health perspective** and the **quality of evidence**. 

2) In so doing you should: **identify gaps in surveillance and control**; and  **make recommendations** for future action.

Code is folded but available via the **Show code** button.

::: {.callout-note}
## Files you need (in your workshop folder)

- `dat_all_anon.csv` — human and animal rabies cases for MIMAROPA
- `mdv_summary.csv` — dog vaccination data (timing of vaccination campaigns in the region by province)
- `rolling_vac_cov_yr.csv` — hupreprocessed estimates of vaccination coverage by municipality
- `PHL_municipality.shp` (and its `.dbf`, `.shx`, `.prj` files) — municipality polygons for MIMAROPA
:::

::: {.callout-note collapse="true"}
## import and process the data for plotting

```{r setup}
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(reactable)
library(scales)

# Set the province names of interest
province_names <- c("Romblon", "Oriental Mindoro", "Marinduque", "Occidental Mindoro")

## load and process the outbreak data
dat_outbreak <- read_csv(here("data","epi","dat_all_anon.csv"))

head(dat_outbreak) # look to identify what columns are in the data

dat_outbreak <- dat_outbreak %>% mutate(date = dmy(date)) # reformat the dates

# Create monthly breaks 
monthly_breaks <- seq.Date(
  from = floor_date(min(dat_outbreak$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak$date), "month"),
  by   = "month"
) - 1 
monthly_breaks  # print out the object - is it creating what you want?

```

:::
---

## Describing the outbreak:

### .1 Epidemic curve 

Monthly reported animal rabies cases (confirmed and probable) and human rabies deaths in Oriental Mindoro, Romblon, Marinduque and Occidental Mindoro are presented in a time series.


```{r epi-curve-overall, fig.width=7, fig.height=4}

## Filter the outbreak data
dat_outbreak_trun <- dat_outbreak %>% 
  filter(date >= "2023-06-30" & date < "2024-12-31") # Work out how to show different time windows

# now create monthly breaks for a truncated period
monthly_breaks_trun <- seq.Date(
  from = floor_date(min(dat_outbreak_trun$date, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak_trun$date), "month"),
  by   = "month"
) - 1
 
# Set the colour palettes used for plots
pal_status <- c("Human Death"        = "#A3003F",
                "Probable (Animal)"  = "#E59500",
                "Confirmed (Animal)" = "#3C8590")


# Now create a timeseries of these data
ggplot(dat_outbreak_trun, # indicate the data set you plan to use
       aes(x = date, fill = case_status_fct)) + # set the x-axis and fill colour for the stacked bars
  geom_histogram(breaks = monthly_breaks_trun, col = "white") + # format the histogram with monthly breaks and white borders
  scale_x_date(expand = c(0, 0), date_breaks = "3 months", date_labels = "%b %Y") + # format the x-axis - show every 3 months and label month and year
  scale_fill_manual(values = pal_status) + # set the fill colours for the different case types
  labs(y = "Number of cases", x = "", fill = "") + # add labels for the y-axis and legend
  theme_bw() + # set the theme to black and white
  theme(axis.text.x  = element_text(angle = 40, vjust = 0.5), # format the x-axis text to be angled and vertically justified
        legend.position = "top") # set the legend position to the top of the plot
```

### .2 Locations of cases

The spatial distribution of all these cases are mapped. Human deaths are 
shown as triangles, whilst animal cases are shown as circles.

```{r map, fig.width=6, fig.height=6}

## Geographic data for mapping
municipality <- st_read(here("data", "gis"), layer = "PHL_municipality", quiet = TRUE)
village_loc  <- read.csv(here("data/gis", "PHL_village_centroids.csv"))

# Attach coordinates
village_loc_filt <- village_loc %>% # filter the village_loc data to only include the provinces of interest
  filter(str_starts(Loc_ID, paste(paste0(province_names, "-"), collapse = "|"))) 
# this is quite a complex operation! you can explore how it works later, but basically it is filtering the Loc_ID column to only include those that start with the province names of interest

# Now join the coordinates to the outbreak data
dat_mapped <- dat_outbreak_trun %>% # create a new dataframe (dat_mapped) that contains the outbreak data and coordinates
  mutate(loc_id = paste(province, municipality, barangay, sep = "-")) %>% # create a column (loc_id) that combines the province, municipality and barangay columns
  left_join(select(village_loc_filt, Loc_ID, Latitude, Longitude), 
            by = c("loc_id" = "Loc_ID")) # join the datasets (village_loc_filt, & the outbreak data) using the loc_id column
            
municipality_df <- municipality %>%
  filter(NAME_1 %in% province_names) # filter the municipality data to only include the provinces of interest

# Set the colour palette for the case types
pal_type   <- c("Human" = "#A3003F",
                "Animal" = "#3C8590")

# Now create the map
ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) + # municipality boundaries in grey outline
  geom_point( 
    data = dat_mapped, # use the dat_mapped dataframe for the points
    aes(x = Longitude, y = Latitude, # set the x and y coordinates for the points (human and animal cases)
        colour = case_type, shape = case_type, alpha = case_type), # set the colour, shape and transparency of the points based on the case type
    position = position_jitter(h = 0.01, w = 0.01), # add some jitter to the points to avoid overlap
    size = 2 # set the size of the points
  ) +
  scale_colour_manual(values = pal_type) + #  
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) + # set the point shape for human & animal cases (triangle, circle)
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.6)) + # set the transparency for points (human = 1, animal = 0.6)
  labs(colour = "", shape = "", alpha = "") + 
  # annotate provinces manually as the labels are not in the shapefile
  annotate("text", x = c(122.4,122.4,121.3,120.5), 
           y = c(12.3,13.4,13.6,12.7), 
           label = c("Romblon","Marinduque","Oriental Mindoro","Occidental Mindoro"), 
           size = c(3,3,3,3)) +
  theme_minimal(base_size = 13) + 
  theme(legend.position  = c(0.85, 0.85),
        panel.grid.major = element_blank(),
        axis.text        = element_blank(),
        axis.title       = element_blank())
```


---

## How did the epidemiological situation evolve in each province?

Looking at the epidemic curve by province reveals whether provinces were affected
simultaneously or sequentially — this is important for understanding routes of disease spread and the timing of introduction events

### Oriental Mindoro

> **Key questions and things to explore:**
>
> - What is happening in the early outbreak period? 
> - How do animal cases compare to human cases in Oriental Mindoro?
> - Roughly 1 person is bitten every 3 rabid dogs. Of people bitten, 90% get PEP. Of the 10% that don't get PEP, 20% will develop rabies and die. How does this change your interpretation of the early outbreak period?
> - How does the most recent human death in April 2026 (look at the 2026 map!) change your interpretation of what is happening in Oriental Mindoro? 

```{r epi-curve-ormin, fig.width=9, fig.height=3.5}

# Import the mass dog vaccination data to include in the plots
mdv_summary <- read.csv(here("data", "epi", "mdv_summary.csv"))
mdv_rolling <- read_csv(here("data", "epi", "rolling_vac_cov_yr.csv"))

# Filter MDV data for Oriental Mindoro and create a date column
mdv_ormin <- mdv_summary %>%
  filter(region == "Ormin") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

# Filter outbreak data for Oriental Mindoro
dat_outbreak %>%
  filter(province == "Oriental Mindoro") %>%
  ggplot(aes(x = date, fill = case_status_fct)) + # stacked barplot with date on x-axis and fill colour based on case status
  # add rectangles to indicate the timing of mass dog vaccination campaigns
  geom_rect( 
    data = mdv_ormin,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  # add histogram of cases with monthly breaks and white borders
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(
    values = c(pal_status)
  ) +
  # add labels for the y-axis, x-axis and legend
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 40, vjust = 0.5),
    legend.position  = "top",
    strip.background = element_rect(fill = "#f0f0f0"),
    strip.text       = element_text(face = "bold")
  )

```
*Dates of mass dog vaccination campaigns are shown in grey (a campaign is defined as >700 dogs vaccinated in a municipality in a month)*

#### How do human cases compare to animal cases in Oriental Mindoro?

```{r epi-curve-facet-ormin, fig.width=9, fig.height=5}

# Filter outbreak data for Oriental Mindoro and create a histogram of cases with monthly breaks and white borders, 
# facet by case type (human or animal)
dat_outbreak %>% 
  filter(province == "Oriental Mindoro") %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~case_type, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

#### Does the distribution of cases over time and across municipalities suggest any gaps in surveillance (from the health or animal health perspective)?

```{r map-ormin, fig.width=13, fig.height=13}

# Attach coordinates to outbreak data for Oriental Mindoro
dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Oriental Mindoro") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

# Filter municipality data for Oriental Mindoro and join with vaccination coverage data
municipality_df <- municipality %>%
  filter(NAME_1 == "Oriental Mindoro") %>% 
  left_join(mdv_rolling, by = c("NAME_2" = "municipality")) %>% 
  filter(NAME_2 != "Naujan Lake") ## Removing water body which was causing NA

# Now create the map of cases and vaccination coverage
ggplot() +
  # add municipality polygons with fill colour based on vaccination coverage
  geom_sf(data = municipality_df, aes(fill = vacc_cov_roll * 100), alpha = 80) +
      scale_fill_gradient(
      name   = "Vaccination \ncoverage",
      low    = "#FFFFFB", high = "#DFA801",
      labels = comma, limits = c(0, 100)
    ) +
  # add points for human and animal cases with jitter to avoid overlap
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year,ncol=4) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```
*Note: some vaccination has occurred in 2026, however this data has not yet been released*


### Romblon

> **Key questions and things to consider:**
>
> - Romblon province is made up of different islands. How did rabies spread across the islands over time?
> - If you were facing the situation in Tablas island in December 2022 (look at the 2022 map), what actions would you recommend to the local authorities?
- When did the first death occur? What can you say about the responsiveness of dog vaccination efforts? 
> - What if you were facing the situation in Sibuyan island in December 2023, would you be concerned? what communications would you want between municipalities? How might your thinking change by August 2024? Do you have any concerns if you are from Corcuera island?
>  What do these complete plots tell you about the current surveillance efforts in Romblon?


#### How does the outbreak look in Romblon if we break this down by island?

```{r epi-curve-romblon, fig.width=9, fig.height=5}

# Filter outbreak data for Romblon and create a histogram of cases with monthly breaks and white borders, facet by island
dat_romblon = dat_outbreak %>% 
  filter(province == "Romblon")

# Create monthly breaks for Romblon outbreak data
monthly_breaks <- seq.Date(
  from = floor_date(min(dat_romblon$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_romblon$date), "month"),
  by   = "month"
) - 1

# Filter MDV data for Romblon and create date columns for the start and end of each campaign
mdv_romblon <- mdv_summary %>%
  filter(region == "Romblon") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

# Now create the epidemic curve for Romblon with rectangles indicating the timing of mass dog vaccination campaigns
dat_romblon %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  # add rectangles to indicate the timing of mass dog vaccination campaigns
    geom_rect(
    data = mdv_romblon,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  # add histogram of cases with monthly breaks and white borders
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```
*Dates of mass dog vaccination campaigns are shown in grey (MDV campaign defined as >700 dogs vaccinated in a municipality in a month)*


```{r epi-curve-island, fig.width=9, fig.height=7}
# Now create the epidemic curve for Romblon with rectangles indicating the timing of mass dog vaccination campaigns, 
# facet by island
dat_romblon %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~island, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

#### Spatial distributions of human and animal rabies cases in Romblon by year

```{r map-romblon, fig.width=13, fig.height=11}
# Attach coordinates to outbreak data for Romblon
dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Romblon") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

# Filter municipality data for Romblon and join with vaccination coverage data
municipality_df <- municipality %>%
  filter(NAME_1 == "Romblon") %>% 
  left_join(mdv_rolling, by = c("NAME_2" = "municipality")) %>% 
  filter(year >= 2022) ## filtering as no cases in Romblon until 2022
  
# Now create the map of cases and vaccination coverage for Romblon
ggplot() +
  geom_sf(data = municipality_df, aes(fill = vacc_cov_roll * 100), alpha = 80) +
      scale_fill_gradient(
      name   = "Vaccination \ncoverage",
      low    = "#FFFFFB", high = "#DFA801",
      labels = comma, limits = c(0, 100)
    ) +
  # add points for human and animal cases with jitter to avoid overlap
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year) +
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```
*Note: some vaccination has occurred in 2026, however this data has not yet been released*

### Marinduque

> **Key questions and things to consider:**
>
> - How did the outbreak evolve in Marinduque over time?
> - What do these plots tell you about the current surveillance efforts in Marinduque?
> - When did the first death occur? And how does this correspond to the increase in dog vaccination efforts? 
> - How does the outbreak in Marinduque compare to the outbreaks in Oriental mindoro and Romblon provinces?
> - How does the human death in Dec 2025 impact your interpretation?


```{r epi-curve-marin, fig.width=9, fig.height=4}

# Filter MDV data for Marinduque and create date column
mdv_marin <- mdv_summary %>%
  filter(region == "Marinduque") %>%
  mutate(
    date_start = dmy(paste("01", month, year)),
    date_end   = date_start + days(30)
  )

# Filter outbreak data for Marinduque
dat_marin = dat_outbreak %>% 
  filter(province == "Marinduque")

# Create monthly breaks for Marinduque outbreak data
monthly_breaks <- seq.Date(
  from = floor_date(min(dat_marin$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_marin$date), "month"),
  by   = "month"
) - 1

# Now create the epidemic curve for Marinduque with rectangles indicating the timing of mass dog vaccination campaigns
dat_outbreak %>% 
  filter(province == "Marinduque") %>% 
ggplot(aes(x = date, fill = case_status_fct)) +
  # add rectangles to indicate the timing of mass dog vaccination campaigns
  geom_rect(
    data = mdv_marin,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = Inf, fill = "grey70"),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  # add histogram of cases with monthly breaks and white borders
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~island, ncol = 1) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```
*Dates of mass dog vaccination campaigns are shown in grey (a campaign defined as >700 dogs vaccinated in a municipality in a month)*



```{r map-marin, fig.width=13, fig.height=11}

# Attach coordinates to outbreak data for Marinduque
dat_mapped <- dat_outbreak %>%
  mutate(
    loc_id = paste(province, municipality, barangay, sep = "-"),
    year   = lubridate::year(date)
  ) %>%
  filter(province == "Marinduque") %>%
  left_join(
    select(village_loc, Loc_ID, Latitude, Longitude),
    by = c("loc_id" = "Loc_ID")
  )

# Filter municipality data for Marinduque and join with vaccination coverage data
municipality_df <- municipality %>%
  filter(NAME_1 == "Marinduque") %>% 
  left_join(mdv_rolling, by = c("NAME_2" = "municipality")) %>% 
  filter(year >= 2023) ## filtering as no cases in Marinduque until 2022
  
# Now create the map of cases and vaccination coverage for Marinduque
ggplot() +
  # add municipality polygons with fill colour based on vaccination coverage
  geom_sf(data = municipality_df, aes(fill = vacc_cov_roll * 100), alpha = 80) +
      scale_fill_gradient(
      name   = "Vaccination \ncoverage",
      low    = "#FFFFFB", high = "#DFA801",
      labels = comma, limits = c(0, 100)
    ) +
  # add points for human and animal cases with jitter to avoid overlap
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.8)) +
  facet_wrap(~ year, ncol = 2) + # make a map for every year and arrange them in 2 columns
  labs(
    colour = "", shape = "", alpha = "" ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.major = element_blank(),
    axis.text        = element_blank(),
    axis.title       = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```
*Note: some vaccination has occurred in 2026, however this data has not yet been released*

---

## Providing the regional context

> **Key questions and things to explore:**
>
> - Which province do you think has the most robust surveillance? (consider surveillance from the veterinary perspective, the public health perspective and under a One Health approach) 
> - Do you consider rabies to be endemic in any of the provinces? 
> - What do you think are the triggers of outbreak responses?
> - Why do the ratios of human to animal cases differ in the three provinces?

### Extended epidemic curve 

Monthly reported animal cases (confirmed and probable) and human deaths in Oriental Mindoro, Romblon, Marinduque and Occidental Mindoro including all data from 2019 - 2026

```{r epi-curve-region, fig.width=9, fig.height=3.5}
# Create monthly breaks for the full outbreak data
monthly_breaks <- seq.Date(
  from = floor_date(min(dat_outbreak$date - 60, na.rm = TRUE), "month"),
  to   = ceiling_date(max(dat_outbreak$date), "month"),
  by   = "month"
) - 1

# Now create the epidemic curve for the full outbreak data with rectangles indicating the timing of mass dog vaccination campaigns
ggplot(dat_outbreak, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "3 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  labs(y = "Number of cases", x = "", fill = "") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top")
```

### Epidemic curve by province

```{r epi-curve-facet, fig.width=9, fig.height=8}
# Now create the epidemic curve for the full outbreak data with rectangles indicating the timing of mass dog vaccination campaigns, facet by province
ggplot(dat_outbreak, aes(x = date, fill = case_status_fct)) +
  geom_histogram(breaks = monthly_breaks, col = "white") +
  scale_x_date(expand = c(0, 0), date_breaks = "4 months", date_labels = "%b %Y") +
  scale_fill_manual(values = pal_status) +
  facet_wrap(~province, ncol = 1) + # facet by province and arrange in 1 column
  labs(y = "Number of cases", x = "", fill = "",
       title = "Reported rabies cases by province") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 40, vjust = 0.5),
        legend.position = "top",
        strip.background = element_rect(fill = "#f0f0f0"),
        strip.text       = element_text(face = "bold"))
```

### Regional location of cases

Spatial distribution of all cases from 2019 to 2026. Human deaths are
shown as triangles, whilst animal cases are shown as circles.

```{r map-region, fig.width=9, fig.height=9}

# Attach coordinates
village_loc_filt <- village_loc %>%
  filter(str_starts(Loc_ID, paste(paste0(province_names, "-"), collapse = "|")))

# Now join the coordinates to the outbreak data
dat_mapped <- dat_outbreak %>%
  mutate(loc_id = paste(province, municipality, barangay, sep = "-")) %>%
  left_join(select(village_loc_filt, Loc_ID, Latitude, Longitude),
            by = c("loc_id" = "Loc_ID"))

# Filter municipality data for the provinces of interest
municipality_df <- municipality %>%
  filter(NAME_1 %in% province_names) 

# Now create the map of cases for the provinces of interest
ggplot() +
  geom_sf(data = municipality_df, alpha = 0, colour = "dimgrey", linewidth = 0.4) +
  # add points for human and animal cases with jitter to avoid overlap
  geom_point(
    data = dat_mapped,
    aes(x = Longitude, y = Latitude,
        colour = case_type, shape = case_type, alpha = case_type),
    position = position_jitter(h = 0.01, w = 0.01),
    size = 2
  ) +
    facet_wrap(~ year) + # make a map for every year
  scale_colour_manual(values = pal_type) +
  scale_shape_manual(values  = c("Human" = 17, "Animal" = 16)) +
  scale_alpha_manual(values  = c("Human" = 1,  "Animal" = 0.7)) +
  labs(colour = "", shape = "", alpha = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        panel.grid.major = element_blank(),
        axis.text        = element_blank(),
        axis.title       = element_blank())

```

### How do human deaths compare with animal cases?

A critical question is whether surveillance is sufficient. If we have areas where human deaths occur 
with relatively few reported animal cases this would suggest under-reporting
of animal cases. Therefore, the annual incidence and the ratio of human deaths to animal cases may provide an indication of the surveillance sensitivity and reporting.


#### Ratio of human deaths to animal cases

How do the ratios of human deaths to animal cases vary across provinces? 
What do these suggest about the surveillance strength?

```{r ratio-table}
dat_outbreak %>%
  group_by(province) %>%
  summarise(
    Animal_cases  = sum(case_type == "Animal"),
    Human_deaths  = sum(case_type == "Human"),
    .groups = "drop"
  ) %>%
  mutate(Deaths_per_100_animal = round(Human_deaths / Animal_cases * 100, 1)) %>%
  rename(Province               = province,
         `Animal cases`         = Animal_cases,
         `Human deaths`         = Human_deaths,
         `Deaths per 100 animal cases` = Deaths_per_100_animal) %>%
  reactable(striped = TRUE, highlight = TRUE, pagination = FALSE)
```


### Is rabies an emerging infectious disease threat or a neglected disease of poverty?
Rabies can be viewed as an emerging disease that can cause outbreaks in areas where it was previously absent. But it is also a disease that is entirely vaccine-preventable, that can be eliminated through dog vaccination, whilst human deaths only occur if people at-risk do not access emergency life-saving vaccines after an exposure.

The last decade in MIMAROPA region illustrates all of these different situations. Can you quantify the incidence of human rabies deaths and of rabies in dogs in each province and how this varies by year?
What is the highest incidence? and what is the lowest incidence?
Taking these extremes, and extrapolating to the entirety of the Philippines, what is the worst case scenario? how many rabies deaths *could* occur across the country if they were all to face the same challenges in rabies control and prevention?

And what if rabies control improved and surveillance was better - what is the best case scenario? 
What are the obstacles to achieving this best case outcome? and what can be done better that can support the best outcome?!

#### Population estimates by province

| Location | Human population | Dog population |
|---|--:|--:|
| Oriental Mindoro | 900,000 | 200,000 |
| Occidental Mindoro | 510,000 | 110,000 |
| Marinduque | 220,000 | 40,000 |
| Romblon | 284,000 | 55,800 |


#### Incidence of human deaths and animal cases

Calculating the annual incidence of human deaths and animal cases by province

```{r incidence-table}
# Population data by province
dog_pop <- c(
  "Oriental Mindoro"  = 200000,
  "Occidental Mindoro" = 110000,
  "Marinduque"        = 40000,
  "Romblon"           = 55800   # 30000 + 15000 + 2000 + 8800
)

human_pop <- c(
  "Oriental Mindoro"  = 920000,
  "Occidental Mindoro" = 510000,
  "Marinduque"        = 220000,
  "Romblon"           = 284000  # 170000 + 65000 + 9000 + 40000
)

dat_outbreak <- dat_outbreak %>%
  mutate(years = as.numeric(max(date, na.rm = TRUE) - 
                            min(date, na.rm = TRUE)) / 365.25) 

# Summary by province
province_summary <- dat_outbreak %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(province) %>%
  summarise(
    animal_cases = sum(case_type == "Animal"),
    human_deaths = sum(case_type == "Human"),
    years        = first(years),
    dogs         = dog_pop[first(province)],
    humans       = human_pop[first(province)],
    .groups = "drop"
  ) %>%
  mutate(
    animal_incidence = round(animal_cases / (dogs * years) * 1000, 2),
    human_incidence  = round(human_deaths / (humans * years) * 100000, 2)
  )

# Overall row
overall_row <- dat_outbreak %>%
  summarise(
    province     = "Overall",
    animal_cases = sum(case_type == "Animal"),
    human_deaths = sum(case_type == "Human"),
    years        = first(years),
    dogs         = sum(unique(dog_pop)),
    humans       = sum(unique(human_pop)),
    animal_incidence = round(animal_cases / (dogs * years) * 1000, 2),
    human_incidence  = round(human_deaths / (humans * years) * 100000, 2)
  )

# Worst case: highest human incidence in any single year + province
worst_case <- dat_outbreak %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(province, year) %>%
  summarise(
    human_deaths = sum(case_type == "Human"),
    .groups = "drop"
  ) %>%
  mutate(
    humans = human_pop[province],
    human_incidence = round(human_deaths / humans * 100000, 2)
  ) %>%
  slice_max(human_incidence, n = 1)


# Bind and display
tbl <- bind_rows(province_summary, overall_row)

tbl %>%
  select(
    Province                                      = province,
    `Animal cases`                                = animal_cases,
    `Animal incidence (per 1,000 dogs/year)`      = animal_incidence,
    `Human deaths`                                = human_deaths,
    `Human incidence (per 100,000 people/year)`   = human_incidence
  ) %>%
  reactable(
    striped = TRUE,
    highlight = TRUE,
    pagination = FALSE,
    rowStyle = function(index) {
      if (tbl$province[index] == "Overall") {
        list(fontWeight = "bold")
      }
    }
  )


```


#### How does the annual incidence extrapolate? 

Romblon in 2023 and 2025 recorded the highest annual human rabies incidence at 1.06 deaths per 100,000 people. 
What happens if we extrapolate these numbers to the overall population 
(population estimate of the Philippines ~113 million)?


```{r death-scenario}
total_pop = 113000000
av_inc = 0.38
wost_inc = 1.06

tibble(
  Scenario        = c("Regional average", "Worst case"),
  Rate            = c(av_inc, wost_inc),
  Expected_deaths = c(round(total_pop * (av_inc / 100000),0), 
                      round(total_pop * (wost_inc / 100000),0))
) %>%
  reactable(
    striped    = TRUE,
    highlight  = TRUE,
    pagination = FALSE,
    columns = list(
      Scenario = colDef(name = "Incidence"),
      Rate = colDef(
        name   = "Rate (per 100,000/year)",
        format = colFormat(digits = 2)
      ),
      Expected_deaths = colDef(
        name   = "Expected deaths/year (Philippines)",
        format = colFormat(separators = TRUE)
      )
    )
  )

```




